---
title: "Final Project-Heart Disease"
author: "Osheen Arya"
output: html_notebook
date: "Due Date:  12/13/2020"
---

#Introduction 
One of the courses I took this semester was Kinesiology 81: Wellness Theory and in this course I learned a lot about heart disease and its causes. This sparked my interest in the topic and for my final project for STAT 184 I will be exploring it further using R skills I learned in this course. 

I will be using a examining two data sets for this project. The first source is a dataset from Kaggle (HeartDisease.csv) containing information about the presence of heart disease in patients. This data set is very informative as usefull by itself because it contains many attributes that can indicate risk factors and causes of heart disease. 

The second data set (FraminghamStudy.csv) is from the Framingham Heart Disease Study which is a epidimeology study that began in 1948. The objectives of the Framingham Study are to study the incidence and prevalence of cardiovascular disease (CVD) and its risk factors, trends in CVD incidence and its risk factors over time, and familial patterns of CVD and risk factors. This data set is incredibly helpful as it includes information on many different risk factor for CVD. 

#Set-Up
##Clean Environment and Load Necessary Packages 
```{r}
# clean up the RStudio environment 
rm(list = ls())

# load packages
install.packages("party")
library(mosaic)
library(tidyverse)
library(tidyr)
library(rvest)
library(lubridate)
library(ggplot2)
library(readr)
```
##Guiding Question 
**Do more males or females have heart disease and does high cholestrol lead to heart disease? **

##Load Data

###Heart Disease Data
```{r}
HeartDisease <- read_csv("HeartDisease.csv")
View(HeartDisease)
```

```{r}
#The First 5 Rows of the Data Set
head(HeartDisease, 5)
```

Each case in this data set has information on each patient. The last column "target" indicates the presence of heart disease in the patient. It is integer valued 0 = no disease and 1 = disease.

####Dimensions, number of rows and columns in the data set. 
```{r}
dim(HeartDisease)
nrow(HeartDisease)
ncol(HeartDisease)
```

####Understanding the datatype of the data set
```{r}
str(HeartDisease)
```

####Summary of the Data Set 
```{r}
summary(HeartDisease)
```

###Framingham Study Data 
```{r}
FraminghamStudy <- read_csv("FraminghamStudy.csv")
View(FraminghamStudy)
```
```{r}
#The First 5 Rows of the Data Set
head(FraminghamStudy, 5)
```
Each case shows data on each person observed in the study and gives additional information such as BMI, glucose and the data about their smoking habits.

####Dimensions, number of rows and columns in the data set. 
```{r}
dim(FraminghamStudy)
nrow(FraminghamStudy)
ncol(FraminghamStudy)
```

####Understanding the datatype of the data set
```{r}
str(FraminghamStudy)
```

####Summary of the Data Set 
```{r}
summary(FraminghamStudy)
```

#Key Visualizations 

##Description Of Variables (HeartDisease Data)
age - age in years
sex - sex (1 = male; 0 = female)
cp - chest pain type (1 = typical angina; 2 = atypical angina; 3 = non-anginal pain; 4 = asymptomatic)
trestbps - resting blood pressure (in mm Hg on admission to the hospital)
chol - serum cholestoral in mg/dl
fbs - fasting blood sugar > 120 mg/dl (1 = true; 0 = false)
restecg - resting electrocardiographic results (0 = normal; 1 = having ST-T; 2 = hypertrophy)
thalach - maximum heart rate achieved
exang - exercise induced angina (1 = yes; 0 = no)
oldpeak - ST depression induced by exercise relative to rest
slope - the slope of the peak exercise ST segment (1 = upsloping; 2 = flat; 3 = downsloping)
ca - number of major vessels (0-3) colored by flourosopy
thal - 3 = normal; 6 = fixed defect; 7 = reversable defect
target - the predicted attribute - diagnosis of heart disease (angiographic disease status) (Value 0 = < 50% diameter narrowing; Value 1 = > 50% diameter narrowing)

To make is easier to use this data I must convert it from numeric values to its respective type. 

##Converting Data
```{r}
Heart <- HeartDisease %>% 
  mutate(sex = if_else(sex == 1, "MALE", "FEMALE"),
         fbs = if_else(fbs == 1, ">120", "<=120"),
         exang = if_else(exang == 1, "YES" ,"NO"),
         cp = if_else(cp == 0, "TYPICAL ANGINA",
                      if_else(cp == 1, "ATYPICAL ANGINA",
                      if_else(cp == 2,"NON-AGINAL PAIN","ASYPMTOMATIC" ))),
         restecg = if_else(restecg == 0, "NORMAL",
                           if_else(restecg == 1, "ABNORMALITY", "PROBABLE OR DEFINITE")),
         slope = as.factor(slope),
         ca = as.factor(ca),
         thal = as.factor(thal),
         target = if_else(target == 1, "YES", "NO")
         ) %>% 
  dplyr::select(target, sex, fbs, exang, cp, restecg, slope, ca, thal, everything())
```


##Summary after pre-processing the data
```{r}
summary(Heart)
```

##Ensuring Data does not Contain NA values
```{r}
colSums(is.na(Heart))
```

###Number of Females vs Males in the Data Set
```{r}
Heart %>% group_by(sex) %>%
  summarise(count = n())
```

There are more Males than Females in the Patient Data Set.

###Average Cholestrol for Each Sex
```{r}
Heart %>% group_by(sex) %>%
  summarise(avgChol = mean(chol))
```

###Is Patient High or Low Risk 
```{r}
LevelChol <- function(chol){
  if_else(chol <= 200, "Desirable Chol",
                       if_else(chol <= 239,"Borderline high","High risk"))
}
Heart %>% select(sex,age,chol,target) %>%
  mutate(RiskFromChol = LevelChol(chol))
```

Risk level determined for each patient using their cholestrol.

###Average Cholestral Level Among Sex
```{r}
Heart %>% group_by(sex,restecg) %>%
  summarise(avgChol = mean(chol))
```
We have determined the average cholestrol level for each sex depending on their restecg (resting electrocardiographic results). 

###Analysis of "Target" Variable
```{r}
Heart$target <- as.factor(Heart$target)
ggplot(Heart, aes(x=Heart$target, fill=Heart$target)) + 
  geom_bar() +
  xlab("Heart Disease") +
  ylab("Count") +
  ggtitle("Analysis of Presence and Absence of Heart Disease") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absence", "Presence"))+
  facet_grid(.~sex)

```

With this bar graph it is clear to see that if we just look at the target variable in the data set, males have a larger presence of heart disease. 

###Presence of Heart Disease for All Patients 
```{r}
Heart$target <- as.factor(Heart$target)
ggplot(Heart, aes(x=Heart$target, fill=Heart$target)) + 
  geom_bar() +
  xlab("Heart Disease") +
  ylab("Count") +
  ggtitle("Analysis of Presence and Absence of Heart Disease") +
  scale_fill_discrete(name = "Heart Disease", labels = c("Absence", "Presence"))
```
From the above plot it is evident that the number of people with presence of CVD is larger than those with absence. Now I must perform further analysis to see what factors lead to heart disease. 

#Sex vs Cholestrol Analysis
##Cholestrol can have a large effect on you cardiovascular health. When there is too much cholesterol in your blood, it builds up in the walls of your arteries, causing a process called atherosclerosis

```{r}
g<-ggplot(data=Heart)
g+geom_bar(stat="identity",aes(x=as.factor(Heart$sex),y=Heart$chol,fill=Heart$sex))+xlab("Sex")+ylab("cholestrol")+theme(legend.position = "none")
```

Males have higher cholestrol and from our previous plots it clear that they also have a higher presence of CVD.

#Age vs Cholestrol Analysis 
```{r}
g<-ggplot(data=Heart)
g+geom_bar(stat="identity",aes(x=Heart$age,y=Heart$chol,fill=as.factor(Heart$sex)))+xlab("Age")+ylab("cholestrol")+labs(fill="Color Coding")
```

However, when we plot the cholestrol for each sex by their age. Female cholestrol levels are seen as higher. 
```{r}
g<-Heart %>% ggplot(aes(x=age,y=chol,col=sex, size=chol))+geom_point(alpha=0.7)+xlab("Age") + 
           ylab("Cholestoral")+guides(fill = guide_legend(title = "Gender"))
g
```
This plot supports the finding from the bar plot above it. 

#Fasting Blood Sugar Analysis
```{r}
g<-Heart %>% ggplot(aes(x=age,y=chol,col=sex, size=chol))+geom_point(alpha=0.7)+xlab("Age") + 
           ylab("Cholestoral")+facet_grid(.~fbs)
g
```
When we take fast blood sugar into consideration (Test for Type 2 Diabetes). Again we see that or both <= 120 and >120 fbs females have higher cholestrol. 

#Blood Pressure Analysis
```{r}
g<-Heart%>%ggplot(aes(x=sex,y=trestbps))+geom_boxplot(fill="darkorange")+xlab("Sex")+ylab("BP")+facet_grid(~cp)
g
```
#Analysis of Other Factor Using Machine Learning 
```{r}
CompleteCases <- HeartDisease%>%
  select(age,cp,trestbps,chol,fbs,restecg,target) %>%
  na.omit()
CompleteCases %>% nrow()
head(CompleteCases)

```
We need to consider other factors such as age, cp, chol, fbs and restecg to see if they effect presence of heart disease (target).
```{r}
mod1 <- party::ctree(target ~age + cp + chol + fbs + restecg,data=CompleteCases)
plot(mod1, type = "simple")
```

With the result of this plot we will look further into age and cholestrol.

##Age by Cholestral with Smoothening
```{r}
age_point <- ggplot(HeartDisease, aes(age, chol, color = sex, size = chol))+geom_point()+geom_smooth()+theme_classic()+ggtitle("Age by Chol")
age_point
```

#Further Examining Age and Cholestrol with Study Data
```{r}
heart2 <- read.csv("FraminghamStudy.csv")
heart2 <- heart2 %>% select("male","age","totChol")
names(heart2) <- c( "sex", "age","chol")
AgeGroup <- function(age){
  if_else(age <= 30, "20-30",
                       if_else(age <= 40 ,"30-40",
                               if_else(age <= 50,"40-50","60 and above")))
}
set.seed(1)
heart2 <- heart2 %>% 
  mutate(sex = if_else(sex == 1, "MALE", "FEMALE"))
heart2 <- heart2 %>% sample_n(303)
heart2.synthesized <- heart2 %>% mutate(chol =LevelChol(chol)) %>%
  mutate(age = AgeGroup(age)) %>% group_by(age,chol) %>%
  summarise(count = n())
ggplot(data=heart2.synthesized,aes(x=chol,y=count ))+geom_bar(stat='identity',position='stack', width=.9)+ facet_wrap(~age,ncol=3)+ theme(axis.text.x = element_text(angle = 45, hjust=1))
```
#Further Examining Age and Cholestrol with HeartDisease Data 

```{r}
Heart.synthesized <- Heart %>% mutate(chol = LevelChol(chol)) %>%
  mutate(age = AgeGroup(age)) %>% group_by(age,chol) %>%
  summarise(count = n())
ggplot(data=Heart.synthesized,aes(x=chol,y=count ))+geom_bar(stat='identity',position='stack', width=.9)+ facet_wrap(~age,ncol=3)+ theme(axis.text.x = element_text(angle = 45, hjust=1))
```
Using both the data sets we look at was age group is likely to have higher cholestrol. In both the plots it is evident that people age 60 or above and significantly more likely to have a higher cholestrol. 
#Conclusion and Insight on Guiding Question
Looking at the data we were able to say that there is more number of males  suffering from heart diseases than female. We saw that males have higher counts of risk factors that lead to CVD. We were not able to get conclusive evidence on on our subquestion about the effects of high cholestrol. However, we found that poeple age 60 and above are at risk of having high cholestrol. In conclusion, Males who are of age 60 or above are at a greater risk of having high cholestrol which may lead to atherosclerosis. It is evident from our research that they are more likely to have a cardiovascular disease. 

